---
title: "COVID-19 first wave in London, UK"
author: "Emma Davis"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(incidence)
library(EpiEstim)
```

## The Data

First we import the data and assign it to an object, which we call `case_data`. As `read_csv()` is a `tidyverse` function, the object will automatically be of the data type `tibble`, which means we can use other `tidyverse` functions to explore and manipulate the data.

```{r import, echo=FALSE}
case_data <- read_csv('firstwave_London.csv')
```
Then, if we wish to view the raw data, we can simply type the name of the object.

```{r view}
case_data
```

We can see that the data starts with the most recent date at the top of the tibble, but it's more useful to have the data the other way around - with the dates increasing. We can use the function `map_df()` to reverse the direction of the data in the table.

```{r reverse}
case_data <- case_data %>% map_df(rev)
case_data
```

To be able to use the `incidence` package, we now want to convert our data into a new type of object specifically designed for use in this package: an `incidence` object. We do this using the title function of the package, `incidence()`. However, this function is designed to take in a list of dates with each date representing the specimen date of a single case, whereas our data has a single entry for each date, under the column `date`, with total number of cases on that date listed under `newCasesBySpecimenDate`, so we need to do some fiddling to make it work.

First we use the `date` column of our data to generate an incidence object of the correct size (you can refer to a specific column of a data frame or tibble using the `$` symbol). As we have one entry for each date, with no repeats, the function will assume we mean there is one case each day for every date we have given it, so we'll need to fix this before moving on.

## Using the `incidence` package

```{r incidence}
i <- case_data$date %>% incidence()
i
```
We can do this by replacing the `counts` component of our incidence object (also using `$`) with our `newCasesBySpecimenDate` information from our data. We also want to update the `n` component, which represents the total number of cases reported in the data. From here, we can now plot our daily incidence for the time period of our data.

```{r fix}
i$counts <- case_data %>% select(newCasesBySpecimenDate) %>% as.matrix()
i$n <- case_data %>% select(newCasesBySpecimenDate) %>% sum()
plot(i)
```

We can filter by date using two different methods:
1. Using square brackets and numerical index, e.g. `i[1:20]` will select the first twenty dates

```{r dates1}
i_early <- i[1:20]
i_early
plot(i_early)
```

2. Using specific date ranges, e.g. `subset(i, from=as.Date("2020-02-25"))` will select all dates after the 25th Feb 2020.

```{r dates2}
i_mid <- subset(i, from=as.Date("2020-02-25"), to=as.Date("2020-03-10"))
i_mid
plot(i_mid)
```

## Estimating the growth rate and doubling time

Now we have manipulated our data into an appropriate format, and we know how to subset it, we can estimate growth rates and doubling times using the `fit()` function (assuming an exponential growth/decline model), which also gives us the 95% confidence intervals.

```{r fit}
my_fit <- fit(i)
my_fit
```

As well as filtering our dataset, we can also split the dataset into "before" and "after" and get estimates of the growth rate and doubling time for both periods.

```{r split}
split_fit <- fit(i, split=as.Date("2020-03-01"))
split_fit
```

## Using `EpiEstim` to estimate the effective R number

To estimate the R number we need to have some knowledge of the distribution of either the serial interval (onset to onset) or the generation interval (infection to infection). The `EpiEstim` package has a function `estimate_R()` that uses the serial interval and incidence data to estimate the real-time value of the effective R number.

First we need to configure the serial interval distribution settings for this calculation, using the `make_config()` function, by providing a mean and a standard deviation for the serial interval.

```{r serial}
s_interval <- make_config(
  mean_si = 12.0,
  std_si = 5.2
)
```

Next, we use this as an input in the function `estimate_R()`

```{r Re}
my_R <- estimate_R(i,method="parametric_si",config=s_interval)
plot(my_R)
my_R$R
```