---
title: "Week 2 Pair Coding Tasks"
author: "Emma Davis"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(incidence)
library(EpiEstim)
```

## The Data

First we import the data and assign it to an object, which we call `case_data`. I'm using `readRDS()` rather than `read_csv()` because the data in this instance is saved as an RData file (extension .RDS) containing only a vector of dates.

```{r import, echo=FALSE}
case_data <- readRDS('casedates_London.RDS')
```

## Using the `incidence` package

```{r incidence}
i <- case_data %>% incidence()
i
```

We can then look at our data using the `plot()` function.

```{r plot}
plot(i) +
  scale_x_date(date_labels = '%Y')
```

We can filter by date using two different methods: 1. Using square brackets and numerical index, e.g. `i[1:20]` will select the first twenty dates 2. Using specific date ranges, e.g. `subset(i, from = as.Date("2020-02-25"))` will select all dates after the 25th Feb 2020. We can also use both `to =` and `from =` to pick upper and lower bounds.

#### Task 1: Filtering data

-   Create a new incidence object called `firstwave` by filtering to the first four months of the data (up to "2020-06-03").
-   Create another incidence object called `secondwave` by choosing a suitable start and end date (hint: have a look at your incidence plot for the entire range of dates).

```{r Task1}
firstwave <- subset(i, to = '2020-06-03')
firstwave

secondwave <- subset(i, from = '2020-08-03', to = '2021-05-03')
secondwave
```

## Estimating the growth rate and doubling time

We can estimate growth rates and doubling times using the `fit()` function (assuming an exponential growth/decline model), which also gives us the 95% confidence intervals.

#### Task 2: Fitting a model

-   Estimate the growth rate and doubling time for the first half of your `firstwave` object by fitting an exponential growth model
-   Estimate the growth rate and doubling time for the first half of your `secondwave` object by fitting an exponential growth model
-   Use `plot()` to visually assess how good your model is for both cases

```{r Task2}
# First wave
half_firstwave <- subset(firstwave, to = '2020-04-03') # Subset the first half
fit_firstwave <- fit(half_firstwave) # Fit the first half
fit_firstwave # See the fitted growth rate and doubling time
pred_firstwave <- fit_firstwave$info$pred # Extract predictions with uncertainty

# Overlay predictions with incidence data
plot(half_firstwave) +
  geom_line(data = pred_firstwave, aes(x = dates, y = fit)) +
  geom_line(data = pred_firstwave, aes(x = dates, y = lwr),
            linetype = 'dashed') +
  geom_line(data = pred_firstwave, aes(x = dates, y = upr),
            linetype = 'dashed') +
  scale_x_date(date_labels = '%b-%y')

# Second wave
half_secondtwave <- subset(secondwave, to = '2020-12-30') # Subset the 1st half
fit_secondwave <- fit(half_secondtwave) # Fit the first half
fit_secondwave # See the fitted growth rate and doubling time
pred_secondwave <- fit_secondwave$info$pred # Extract predictions w/uncertainty

# Overlay predictions with incidence data
plot(half_secondtwave) +
  geom_line(data = pred_secondwave, aes(x = dates, y = fit)) +
  geom_line(data = pred_secondwave, aes(x = dates, y = lwr),
            linetype = 'dashed') +
  geom_line(data = pred_secondwave, aes(x = dates, y = upr),
            linetype = 'dashed') +
  scale_x_date(date_labels = '%b-%y')
```

As well as filtering our dataset, we can also use the `split = as.Date("...")` argument to split the dataset into "before" and "after" and get estimates of the growth rate and doubling time for both periods. There's also the function `fit_optim_split()`, which will pick the optimal split date for the model.

#### Task 3: Fitting a split model

-   Choose a date you think might be appropriate and use `fit()` and the `split =` argument to get estimates of the growth rate for before and after your chosen date.
-   Use the `fit_optim_split()` function to fit the model and find the optimal date and the growth rates for before and after. How close is the optimal date to your guess?

```{r Task3}
# Let's do it for the second wave!
# Split argument
fit_secondwave_split <- fit(secondwave, split = '2020-12-30')
fit_secondwave_split
pred_secondwave_split_before <- fit_secondwave_split$before$info$pred
pred_secondwave_split_after <- fit_secondwave_split$after$info$pred

plot(secondwave) +
  geom_line(data = pred_secondwave_split_before,
            aes(x = dates, y = fit), colour = 'black') +
  geom_line(data = pred_secondwave_split_before,
            aes(x = dates, y = lwr), linetype = 'dashed', colour = 'black') +
  geom_line(data = pred_secondwave_split_before,
            aes(x = dates, y = upr), linetype = 'dashed', colour = 'black') +
  geom_line(data = pred_secondwave_split_after,
            aes(x = dates, y = fit), colour = 'black') +
  geom_line(data = pred_secondwave_split_after,
            aes(x = dates, y = lwr), linetype = 'dashed', colour = 'black') +
  geom_line(data = pred_secondwave_split_after,
            aes(x = dates, y = upr), linetype = 'dashed', colour = 'black') +
  scale_x_date(date_labels = '%b-%y')
  
# Optimal split (beware of overfitting)
fit_secondwave_optim_split <- fit_optim_split(secondwave)
optim_date <- fit_secondwave_optim_split$split

fit_secondwave_split_opsplit <- fit(secondwave, split = optim_date)
fit_secondwave_split_opsplit
fit_2wave_split_opsplit_before <- fit_secondwave_split_opsplit$before$info$pred
fit_2wave_split_opsplit_after <- fit_secondwave_split_opsplit$after$info$pred

plot(secondwave) +
  geom_line(data = fit_2wave_split_opsplit_before,
            aes(x = dates, y = fit), colour = 'black') +
  geom_line(data = fit_2wave_split_opsplit_before,
            aes(x = dates, y = lwr), linetype = 'dashed', colour = 'black') +
  geom_line(data = fit_2wave_split_opsplit_before,
            aes(x = dates, y = upr), linetype = 'dashed', colour = 'black') +
  geom_line(data = fit_2wave_split_opsplit_after,
            aes(x = dates, y = fit), colour = 'black') +
  geom_line(data = fit_2wave_split_opsplit_after,
            aes(x = dates, y = lwr), linetype = 'dashed', colour = 'black') +
  geom_line(data = fit_2wave_split_opsplit_after,
            aes(x = dates, y = upr), linetype = 'dashed', colour = 'black') +
  scale_x_date(date_labels = '%b-%y')
```

## Using `EpiEstim` to estimate the effective R number

To estimate the R number we need to have some knowledge of the distribution of either the serial interval (onset to onset) or the generation interval (infection to infection). The `EpiEstim` package has a function `estimate_R()` that uses the serial interval and incidence data to estimate the real-time value of the effective R number.

First we need to configure the serial interval distribution settings for this calculation, using the `make_config()` function, by providing a mean and a standard deviation for the serial interval.

```{r serial}
s_interval <- make_config(
  mean_si = 12.0,
  std_si = 5.2
)
```

Next, we use this as an input in the function `estimate_R()`.

#### Example:

```{r Re}
my_R <- estimate_R(i, method = "parametric_si", config = s_interval)
```

#### Task 4: Estimating R

-   Estimate R for your `firstwave` and `secondwave` incidence objects
-   Which date ranges have R greater than 1?
-   Which date ranges have R less than 1?
-   What do you think might have caused these changes in R? Social restrictions

```{r Task4}
r_firstwave <- firstwave |> 
  estimate_R(method = 'parametric_si', config = s_interval)

r_secondwave <- secondwave |> 
  estimate_R(method = 'parametric_si', config = s_interval)
```
