---
title: "Ebola Analysis and Plotting"
author: "Emma Davis"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(outbreaks)
library(incidence)
library(EpiEstim)
```

## Plotting with the `incidence` package

```{r load}
onset <- outbreaks::ebola_sim$linelist$date_of_onset
i <- incidence(onset)
i
```

## Including Plots

You can also embed plots, for example:

```{r simpleplot, echo=FALSE}
plot(i)
```

```{r color, echo=FALSE}
plot(i, color = "forestgreen")
```

```{r transparency, echo=FALSE}
plot(i, alpha=0.8, color="darkred")
```

```{r axislabels, echo=FALSE}
plot(i, xlab="Date case reported", color="gold", alpha=0.5)
```

```{r breaks, echo=FALSE}
plot(i, xlab="Date case reported", n_breaks=10)
```

```{r weekly, echo=FALSE}
i_week <- incidence(onset, interval = 7)
plot(i_week)
plot(i_week, labels_week = FALSE)
```

```{r fit, echo=FALSE}
my_fit <- fit(i)
plot(i, fit = my_fit)
```




```{r splitfit, echo=FALSE}
split_fit <- fit(i, split = as.Date("2014-12-15"))
plot(i, fit = split_fit)
```
The R^2 statistic is used to optimise the average fit of the before and after models

```{r optimfit, echo=FALSE}
optim_split <- fit_optim_split(i)
optim_split
optim_split$split
```

```{r plotoptim, echo=FALSE}
optim_fit <- optim_split$fit
plot(i, fit = optim_fit)
```

```{r neatplot, echo=FALSE}
plot(i, fit = optim_fit, color= "blue", xlab = "Date case reported")
```


## Plotting with the `EpiEstim` package

Reference: WHO Ebola Response Team: Ebola Virus Disease in West Africa - the first 9 months of the epidemic and forward projections. N Engl J Med.

```{r serial}
s_interval <- make_config(
  mean_si = 15.3,
  std_si = 9.3
)
```



```{r Re}
my_R <- estimate_R(i, method = "parametric_si", config = s_interval)
plot(my_R)
```



```{r plots}
plot(my_R, what = "R", legend = TRUE, 
                 options_R = list(col = "red", xlab = "Date"))
```

## Examples using the `ggplot2` package

```{r tibble}
sex <- outbreaks::ebola_sim$linelist$gender
cases <- tibble(onset = onset, sex = sex)
cases
```



```{r histogram}
g <- cases %>% ggplot(aes(x = onset, fill = sex)) + geom_histogram()
g
```



```{r labels}
g <- g + labs(title = "Histogram of onset dates", x = "Date")
g
```



```{r color}
g <- g + scale_fill_manual(values = c("skyblue","darkgreen")) + theme_minimal(base_size = 16)
g
```




```{r legend}
g <- g + scale_fill_manual(values = c("skyblue","darkgreen"), labels = c("female", "male"))
g
```
